Seismic velocities and composition of the Canadian crust

### 2014-04-15 paper/ubcThesisSubmission.tex
thunk

%%
%% This is file `ubcsample.tex',
%% generated with the docstrip utility.
%
% The original source files were:
%
% ubcthesis.dtx  (with options: `ubcsampletex')
%%
%% This file was generated from the ubcthesis package.
%% --------------------------------------------------------------
%% NEED TO FIX: Envelope, ensure abs value --- add to figure caption
%% Lith figure, tighten range.
%% Rework final geochron conclusion. Data is weak for current claims
%% figure -> Figure
\documentclass[msc,oneside]{ubcthesis}

@@

%******** afterpage ***************************
% This package allows you to issue commands at the end of the current
% page.  A good use for this is to use the command
% \afterpage{\clearpage} right after a figure.  This will cause the
% figure to be inserted on the page following the current one (or on
% the current page if it will fit) but will not break the page in the
% middle.
\usepackage{afterpage}

%******** float *********************************
% This package allows you to customize the style of
% "floats"---floating objects such as figures and tables.  In
% addition, it allows you to define additional floating objects which
% may be included in a list similar to that produces by \listoftables
% and \listoffigures.  Common uses include introducing floats for
% programs and other code bits in Compute Science and Chemical Schema.
\usepackage{float}

%******** tocloft *******************************
% This package allows you to customize and define custom lists such
% as a list of programs or Chemical Scheme.  Note: if you use the
% subfigure package, you must specify that you do as an option here.
% The title option uses the default formatting.  We do not use this
% here as the default formatting is acceptable.  Use the float
% package instead unless you need the extra formatting control
% provided by tocloft.
%\usepackage[subfigure, titles]{tocloft}

%******** alltt *********************************
% The alltt package allows you to include files and have them
% formatted in a verbatim fashion.  This is useful for including
% source code from an additional file.
%\usepackage{alltt}

%******** listings ******************************
% The listings package may be used to include chunks of source code
% and has facilities for pretty-printing many languages.
%\usepackage{listings}

%******** longtable *****************************
% The longtable package allows you to define tables that span
% multiple pages.
% \usepackage{longtable}

%******** graphics and graphicx *****************
% This allows you to include encapsulated postscript files.  If you
% don't have this, comment the \includegraphics{} line following the
% comment "%includegraphics" later in this file.
% \usepackage{graphicx}

%******** subfigure *****************************
% The subfigure package allows you to include multiple figures and
% captions within a single figure environment.
%\usepackage{subfigure}

%******** here **********************************
% The here package gives you more control over the placement of
% figures and tables.  In particular, you can specify the placement
% "H" which means "Put the figure here" rather than [h] which means
% "I would suggest that you put the figure here if you think it looks
% good."
%\usepackage{here}

%******** natbib ********************************
% This is a very nice package for bibliographies.  It includes options
% for sorting and compressing bibliographic entries.
\usepackage[numbers,sort&compress]{natbib}

%******** psfrag ******************************
% This allows you to replace text in postscript pictures with formated
% latex text.  This allows you to use math in graph labels
% etc. Uncomment the psfrag lines following the "%psfrag" comment
% later in this file if you don't have this package.  The replacements
% will only be visible in the final postscript file: they will be
% listed in the .dvi file but not performed.
\usepackage{psfrag}

%******** hyperref *****************************
% Please read the manual:
% http://www.tug.org/applications/hyperref/manual.html
%
% This adds hyperlinks to your document: with the right viewers (later
% versions of xdvi, acrobat with pdftex, latex2html etc.) this will
% make your equation, figure, citation references etc. hyperlinks so
% that you can click on them.  Also, your table of contents will be
% able to take you to the appropriate sections.  In the viewers that
% support this, the links often appear with an underscore.  This
% underscore will not appear in printed versions.
%
% Note: if you do not use the hypertex option, then the dvips driver
% may be loaded by default.  This will cause the entries in the list
% of figures and list of tables to be on a single line because dvips
% does not deal with hyperlinks on broken lines properly.
%
% NOTE: HYPERREF is sensitive to the ORDER in which it is LOADED.
% For example, it must be loaded AFTER natbib but BEFORE newly
% defined float environments.  See the README file with the hyperref
% for some help with this.  If you have some very obscure errors, try
% first disabling hyperref.  If that fixes the problem, try various
% orderings.
%
% Note also that there is a bug with versions before 2003/11/30
% v6.74m that cause the float package to not function correctly.
% Please ensure you have a current version of this package.  A
% warning will be issued if you leave the date below but do not have
% a current version installed.
%
% Some notes on options: depending on how you build your files, you
% may need to choose the appropriate option (such as [pdftex]) for the
% backend driver (see the hyperref manual for a complete list).  Also,
% the default here is to make links from the page numbers in the table
% of contents and lists of figures etc.  There are other options:
% excluding the [linktocpage] option will make the entire text a
% hyperref, but for some backends will prevent the text from wrapping
% which can look terrible.  There is a [breaklinks=true] option that
% will be set if the backend supports (dvipdfm for example supports
% it but does not work with psfrag.)
%
% Finally, there are many options for choosing the colours of the
% links.  These will be included by default in future versions but
% you should probably consider changing some now for the electronic
% version of your thesis.
\usepackage[unicode=true,
  linktocpage,
  linkbordercolor={0.5 0.5 1},
  citebordercolor={0.5 1 0.5},
  linkcolor=blue]{hyperref}

% If you would like to compile this sample thesis without the
% hyperref package, then you will need to comment out the previous
% \usepackage command and uncomment the following command which will
% put the URL's in a typewriter font but not link them.
%\newcommand\url[1]{\texttt{#1}}

%******** setspace *******************************
% The setspace package allows you to manually set the spacing of the
% file.  UBC may require 1.5 spacing for microfilming of theses.  In
% this case you may obtain this by including this package and issuing
% one of the following commands:
%\usepackage{setspace}
%\singlespacing
%\onehalfspacing
%\doublespacing

% These commands are optional.  The defaults are shown.  You only
% need to include them if you need a different value
\institution{The University Of British Columbia}

% If you are at the Okanagan campus, then you should specify these
% instead.
%\faculty{The College of Graduate Studies}
%\institutionaddress{Okanagan}
\faculty{The Faculty of Graduate Studies}
\institutionaddress{Vancouver}

% You can issue as many of these as you have...
\previousdegree{B.Sc., The University of British Columbia, 1999}
\previousdegree{M.Sc., The University of British Columbia, 2001}
\previousdegree{Ph.D., Massachusetts Institute of Technology, 2005}

% You can override the option setting here.
% \degreetitle{Jack of All Trades}

% These commands are required.
\title{Seismic velocities and composition of the Canadian crust}
\author{Ben Postlethwaite}
\copyrightyear{2014}
\submitdate{\monthname\ \number\year} % The "\ " is required after
                                      % \monthname to prevent the
                                      % command from eating the space.
\program{GeoPhysics}

% These commands are presently not required for UBC theses as the
% advisor's name and title are not presently required anywhere.
%\advisor{Ariel R.~Zhitnitsky}
%\advisortitle{Professor of Physics}

@@

%% This starts numbering in Roman numerals as required for the thesis
%% style and is mandatory.
\frontmatter

%%% The order of the following components should be preserved.  The order
%%% listed here is the order currently required by FoGS:        \\
%%% Title (Mandatory)                                           \\
%%% Preface (Manditory if any collaborator contributions)       \\
%%% Abstract (Mandatory)                                        \\
%%% List of Contents, Tables, Figures, etc. (As appropriate)    \\
%%% Acknowledgements (Optional)                                 \\
%%% Dedication (Optional)                                       \\

\maketitle                      %% Mandatory
\begin{abstract}                %% Mandatory -  maximum 350 words
  It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. The calibration of models based on evolving mantle fractionation or mantle convection style require accurate estimates of the geological and geophysical properties of crustal provinces to better constrain the details of crustal formation. Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of locally 1-D structure. We also investigate modifications to this approach that can allow discrimination of $V_S$ and $V_P$ under certain conditions. Bootstrap error analysis is performed for each station dataset and subsets of these measurements are compared with results for matching stations from previous studies. Analysis of results in conjunction with additional velocity estimates from active source seismic studies and a seismic property database affords improved constraints on bulk geological composition of the Canadian landmass that are used to evaluate competing models of crustal formation.

  Ben Postlethwaite
  post.ben.here@gmail.com
\end{abstract}

\chapter{Preface} % Manditory if any of the conditions are met

You must include a preface if any part of your research was partly or
wholly published in articles, was part of a collaboration, or required
the approval of UBC Research Ethics Boards.

The Preface must include the following:

\begin{itemize}
\item A statement indicating the relative contributions of all
  collaborators and co-authors of publications (if any), emphasizing
  details of your contribution, and stating the proportion of research
  and writing conducted by you.
\item A list of any publications arising from work presented in the
  dissertation, and the chapter(s) in which the work is located.
\item The name of the particular UBC Research Ethics Board, and the
  Certificate Number(s) of the Ethics Certificate(s) obtained, if
  ethics approval was required for the research.
\end{itemize}

%%% Sections and subsections etc. in the Preface should in general
%%% not be listed in the table of contents, so use the starred form
%%% of \section etc.
\section*{Examples}
Chapter~\ref{cha:apple_ref} is based on work conducted in UBC's Maple
Syrup Laboratory by Dr. A.  Apple, Professor B. Boat, and Michael
McNeil Forbes. I was responsible for tapping the trees in forests X
and Z, conducted and supervised all boiling operations, and performed
frequent quality control tests on the product.

A version of chapter~\ref{cha:apple_ref} has been
published~\cite{Apple:2010}. I conducted all the testing and wrote
most of the manuscript. The section on ``Testing Implements'' was
originally drafted by Boat, B.  Check the first pages of this
chapter to see footnotes with similar information.

Note that this preface must come before the table of contents.  Note
also that this section ``Examples'' should not be listed in the table
of contents, so we have used the starred form: \verb|\section*{Example}|.

\tableofcontents                %% Mandatory
\listoftables                   %% Mandatory if thesis has tables
\listoffigures                  %% Mandatory if thesis has figures

\chapter{Acknowledgements}      %% Optional
This is the place to thank professional colleagues and people who have
given you the most help during the course of your graduate work.

@@

% Chapters are the next main unit.
\chapter{Introduction}

% Sections are a sub-unit
\section{Geology}
The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior comprise the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and southern Superior is composed of roughly east-west trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6 Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011). To the east lies the younger, 1.2-0.98 Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 Ga remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989). To the west of the Superior is the extensive Interior Platform overlying the southern extensions of the Trans-Hudson Orogen, Hearne Domain and Alberta Orogen. The Interior Platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean and Paleoproterozoic crystalline basement. Between the middle of the Jurassic and Paleocene intense metamorphic, plutonic and uplifting events occurred as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coast Mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale d\'{e}collement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

@@

\chapter{Data and Methods}

@@

\section{Teleseismic Data Set}
The primary data utilized in this study are RF processed teleseismic {\it P} wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between 2000 and 2012 and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, POLARIS, FedNor and Chasme. Seismograms are included for analysis if they fall within a 30$^\circ$-100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with sufficiently high signal to noise ratios to permit visual observation of the dominant {\it P} wave energy and with sufficiently impulsive first arrivals to allow arrival onsets to be accurately measured. After selection and filtering more than 80,000 seismograms are available for further processing. Seismic stations contributing to the data utilized in this study are shown in figure \ref{map:stationMap}.

@@

\chapter{Results}

@@

\begin{table}
  \begin{tabular}{ l l l l l l l }
    \cline{2-7}
    & \multicolumn{2}{ c }{ZK} & \multicolumn{3}{ c }{Crust 1.0} & \multicolumn{1}{ c }{Active Source} \\
    \hline
    Region  & $H (km]$& $R$ &$H [km]$& $R$  &$V_P [km/s]$&$V_P [km/s)$ \\
    \hline
    Canada    & 37.0 & 1.75 & 38.1   & 1.74 &   6.41     & 6.33\\
    Shield    & 39.2 & 1.74 & 38.3   & 1.74 &   6.47     & 6.42\\
    Slave     & 38.2 & 1.74 & 38.0   & 1.73 &   6.47     & 6.44\\
    Churchill & 38.9 & 1.73 & 37.9   & 1.74 &   6.46     & 6.39\\
    Superior  & 39.7 & 1.74 & 38.9   & 1.74 &   6.49     & 6.44\\
    Grenville & 41.7 & 1.78 & 39.7   & 1.74 &   6.50     & 6.48\\
    \hline
  \end{tabular}
  \caption[Regional seismic parameter comparsion]{Comparison of seismic velocity and crustal thickness estimates for Canada and selected geological provinces.}
\label{table:regionParameters}

@@

\chapter{Discussion}
\section{Crustal reflectivity}
Seismic refraction and reflection studies of continental crust have afforded valuable constraints on the processes governing crustal evolution. Early efforts to characterize continental crust invoked a 2-layer, slab model comprising lower velocity upper crust and a higher velocity lower crust (Conrad, 1925; Jeffreys, 1926). Evidence supporting this model was originally presented by Conrad (1925) which cites observations of {\it P} arrivals propagating along a mid-crustal interface similar to but shallower than the deeper crust-mantle interface imaged by Andrija Mohorovi\v{c}i\'{c} in 1909 (Litak and Brown, 1989). Interpretations of seismic data invoking both the crustal slab model and the Conrad Discontinuity, as the mid-crustal interface became known, persisted in the literature but were increasingly criticized as being too simplistic, in the case of the slab model, or, lacking consistent evidence (Litak and Brown, 1989) for the Conrad Discontinuity. Technological improvements in reflection seismology allowed researchers to further argue for a complex heterogeneous crust and propose models wherein smooth random lateral inhomogeneities are superimposed on simple velocity gradient models to reproduce the range of seismic observations (Litak and Brown, 1989; Mereu and Ojo, 1981).

@@

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{discInterp.eps}
 \caption[Normalized station depth profiles and SVD analysis]{Left, vertical station depth profiles interpolated to a common Moho. Right, normalized stack of depth profiles, grey line shows mean}
 \label{fig:discInterp}
\end{figure}

@@

We reconcile the teleseismic and active source observations of transparent versus reflective lower crust by considering the scale of vertical heterogeneity to which the two methods are sensitive. Teleseismic body waves include both direct and reflected conversions that can originate from layering with velocity gradients over scale lengths of the order of $\lambda/2$ and $\lambda/4$, respectively, where $\lambda$ is the {\it P}-wavelength (Bostock, 1999). For 1 Hz teleseismic waves, crustal layering with scale lengths of $\sim$ 800-2000 m should be resolvable. Higher frequency active source data (10-40 Hz) employing {\it P}-wave sources are sensitive to finer scale vertical structure with scales of $\sim$ 75-300 m. We infer then that the vertical scales of lower crustal reflectivity are better represented by scale lengths under $\sim$ 800 m yielding a lower crustal structure that is relatively transparent to teleseismic waves but can yield strongly reflective signatures to near-vertical, active-source seismic profiling.  Some authors, for example Mayer et al. (1997) in their study of the Rhine Graben, have employed the term Conrad Discontinuity to refer to an interface or transition between transparent upper crust and reflective lower crust. It is possible in this context that the higher energy levels between 15 and 20 km on the 5th principal component in Figure \ref{fig:disc}, are a manifestation of such signals, but that the lower energy levels ($<5\%$) in this mode would indicate that the Conrad Discontinuity so-defined is at best weakly represented within the Canadian crust. This finding is supported by crustal velocity models characterized by a near constant velocity gradient such as the gradual transition from a more silicic upper crust to a more mafic lower crust in shields and platforms observed by Meissner (1986). It is also supported by the lack of evidence for a consistent mid-crustal reflection boundary in {\sc lithoprobe} data (Hammer et al., 2010). The increased transparency of the lower crust would appear to relate to changes in rheology. Studies of crustal strength (Jackson, 2002), deformation (Burov and Watts, 2006; Meissner et al., 2006) and the inferred $\sim$ 100 m lower crustal lamallae reported by Mooney and Brocher (1987) as well as the widely noted reduction in seismicity of the lower crust *** ADD CITATION *** would support this contention; though see Deichmann (1992), Handy and Brun (2004) and Simpson (1999) for competing evidence.

\section{Crustal Composition}
Mean crustal composition is another important parameter characterizing continental crust though it is important to distinguish it from actual lithological or mineralogical state of the crust. It serves as a representative, average quantity that affords constraint on crustal structure and evolution in space and time. Early attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks (Rudnick and Gao, 2003). After the plate-tectonic revolution, Taylor (1967, 1977), proposed the `island arc' or `andesite' model which posited that bulk crustal chemical composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, average surficial island arc composition is more basaltic than andesitic, and any crustal growth via basaltic basal accretion will only serve to magnify this compositional contrast. Hence this model cannot account for the intermediate composition of post-Archean crust. Recognition of these shortcomings led to the analysis of data from crustal refraction seismic surveys to interrogate middle and lower crustal levels and place constraints on their composition (Christensen and Mooney, 1995; Rudnick and Fountain, 1995). Unlike these latter studies, which also utilize surface and exposed upper-crust geological information, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et al. (1981) although, whereas these authors utilize mean crustal $V_P$, this study combines $V_P/V_S$ from teleseismic measurements with constraints on $V_P$ from previous active source studies as represented in the Crust 1.0 model.

@@

Whereas Zandt and Ammon present evidence for increasing $V_P/V_S$ with crustal age, our estimates signify the opposite trend. Our results are more consistent with those of Durrheim and Mooney (1991) who argue that Proterozoic crust is thicker with higher {\it P}-velocity than Archean crust. The crustal thickness data presented in this study reveals no significant difference between Proterozoic crust (38.9 km) and Archean crust (38.0 km) in contrast to the 5 km difference reported by Durrheim and Mooney. Though the difference is not large, the higher seismic velocity ratio in the Proterozoic units (1.76) versus the Archean units (1.74) aligns with their argument for an evolution of crustal processes throughout geological time. The hypothesis that Proterozoic crust is prone to basaltic underplating, thereby explaining higher velocities and thicker crust, suggested by {\it Durrheim and Mooney} (1991), has been persuasively augmented by recent work (Rudnick 1995; Rudnick and Gao, 2003) to now include a range of possible processes that may explain the data.

\section{Conclusions}

>>>>

\documentclass[msc,oneside]{ubcthesis}

@@

\usepackage{afterpage}

\usepackage{float}

\usepackage[numbers,sort&compress]{natbib}

% version of your thesis.
\usepackage[unicode=true,
  linktocpage,
  linkbordercolor={0.5 0.5 1},
  citebordercolor={0.5 1 0.5},
  linkcolor=blue]{hyperref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\institution{The University Of British Columbia}

\faculty{The Faculty of Graduate Studies}
\institutionaddress{Vancouver}

% You can issue as many of these as you have...
\previousdegree{B.Sc., The University of British Columbia, 2011}

% These commands are required.
\title{Seismic velocities and composition of the Canadian crust}
\author{Ben Postlethwaite}
\copyrightyear{2014}
\submitdate{\monthname\ \number\year} % The "\ " is required after
                                      % \monthname to prevent the
                                      % command from eating the space.
\program{Geophysics}

@@

\frontmatter

\maketitle                      %% Mandatory
\begin{abstract}                %% Mandatory -  maximum 350 words
  It has been suggested that processes driving crustal formation in the Archean and Proterozoic were dissimilar and produced crusts with unique bulk properties and average thicknesses. The calibration of models based on evolving mantle fractionation or mantle convection style require accurate estimates of the geological and geophysical properties of crustal provinces to better constrain the details of crustal formation. Fifteen years of publicly accessible teleseismic data from all available Canadian seismic stations are binned in horizontal slowness and deconvolved into receiver functions. We apply a stacking method to retrieve estimates of the bulk crustal velocity ratio $V_P/V_S$ and thickness $H$ from these data under the assumption of locally 1-D structure. We also investigate modifications to this approach that can allow discrimination of $V_S$ and $V_P$ under certain conditions. Bootstrap error analysis is performed for each station dataset and subsets of these measurements are compared with results for matching stations from previous studies. Analysis of results in conjunction with additional velocity estimates from active source seismic studies and a seismic property database affords improved constraints on bulk geological composition of the Canadian landmass that are used to evaluate competing models of crustal formation.
\end{abstract}

\chapter{Preface}

Teleseismic data discussed in chapter \ref{cha:datamethods} section \ref{section:seismicdata} was processed using Python scripts written by myself and a set of shell scripts based off previous work by Michael Bostock. Deconvolution algorithms and coordinate transformation algorithms are likewise interpreted from previous work by Michael Bostock. Suggestions by David Snyder to investigate signals of the Conrad Discontinuity lead to the analysis of crustal discontinuities in chapter \ref{cha:results} and chapter \ref{cha:discussion}. Nik Christensen provided invaluable support during the discussion on crustal composition in chapter \ref{cha:discussion} section \ref{section:crustalcomp}. Specifically, Nik providing assistance in the interpretation and use of laboratory velocity measurements (Christensen, 1996) and rock chemical analyses (Le Maitre, 1976). His suggestions on crustal compositional literature and recommendations for the layout of figure \ref{fig:lith} were critical to this discussion. Michael Bostock provided advice and editorial recommendations in all chapters of this thesis.

\tableofcontents
\listoftables
\listoffigures

@@

% Chapters are the next main unit.
\chapter{Introduction} \label{cha:intro}

% Sections are a sub-unit
\section{Geology}
The Canadian continental lithosphere is composed of at least fifteen large geological provinces representing crustal evolution from the early Archean to the present accretionary growth and orogenic deformation in the Western Cordillera. The Precambrian Canadian Shield and the bounding platforms of North America's interior comprise the cratonic core of Laurentia (Hammer et al., 2010). Several distinct geological provinces brought together during a series of Paleoproterozoic orogens make up the Canadian Shield (Hoffman, 1988, Figure 1). The largest of these provinces, and the world's largest Archean craton, is the Superior Province. Bounded by Proterozoic orogenic belts, the western and southern Superior is composed of roughly east-west trending geological subprovinces that include the granite-greenstone terranes of the Abitibi Greenstone Belt, metasedimentary gneisses, high-grade gneisses and plutonic complexes assembled during the period between 3-2.6 Ga (Thurston et al., 1991). In the northwest of the Superior Province on the eastern shores of the Hudson Bay lies the Nuvvuagittuq greenstone belt that recent geochronological work has dated at ~4.28 Ga making it the only known remnant of Hadean crust preserved on Earth (O'Neil et al., 2011).

To the east lies the younger, 1.2-0.98 Ga, Grenville orogen, consisting largely of Superior and Southern Province rock reworked during Alpine-Himalayan like continental collision (Eaton et al., 2006). North of the Superior Province and separated by the Paleoproterozoic Trans-Hudson Orogen are the Rae and Hearne domains which together form the western extent of the Churchill Province. Studies of the Rae suggest a Paleoarchean and Mesoarchean basement whereas the Hearne comprises granite-and-greenstone terranes of ages ~2.7Ga (Thompson et al., 2010). The Slave Province is a relatively small Archean granite-greenstone block lying to the northwest of the Churchill province. Its western portion is the 4.0 Ga remnant of a larger Archean continent, whereas the eastern Slave forms a younger (2.6 Ga) paired accretionary prism and island arc (Kusky, 1989).

To the west of the Superior is the extensive Interior Platform overlying the southern extensions of the Trans-Hudson Orogen, Hearne Domain and Alberta Orogen. The Interior Platform is largely Archean reworked during the Paleoproterozoic with bore-hole samples indicating amphibolite to granulite-grade Proterozoic metamorphism (Hammer et al., 2010). Extending west from the Interior platform, the Western Cordillera developed in a passive margin setting on and adjacent to late Archean and Paleoproterozoic crystalline basement. Between the middle of the Jurassic and Paleocene intense metamorphic, plutonic and uplifting events occurred as the result of Paleozoic and younger intraoceanic-arc accretionary events. The Coast Mountains evolved during the late Cretaceous as a wide magmatic arc composed largely of granitic rock (Clowes et al, 1995). Although there is considerable variation in structure and tectonic history, roughly half of the Cordillera is underlain by Proterozoic-margin crust that forms a crustal-scale d\'{e}collement beneath interleaved and duplexed crust (Hammer et al., 2010). The rich tectonic tapestry of the Canadian continental crust affords an excellent opportunity to study variations in bulk crustal properties with geological age.

@@

\chapter{Data and Methods} \label{cha:datamethods}

@@

\section{Teleseismic Data Set} \label{section:seismicdata}
The primary data utilized in this study are RF processed teleseismic {\it P} wave seismograms originating from more than 700 earthquake sources. These earthquakes occurred between 2000 and 2012 and were recorded on subsets of 343 broadband seismic stations distributed across Canada. Seismic stations are selected from all available regional and national networks including CNSN, POLARIS, FedNor and Chasme. Seismograms are included for analysis if they fall within a 30$^\circ$-100$^\circ$ epicentral distance window. Quality control is performed by admitting only those seismograms with sufficiently high signal to noise ratios to permit visual observation of the dominant {\it P} wave energy and with sufficiently impulsive first arrivals to allow arrival onsets to be accurately measured. After selection and filtering more than 80,000 seismograms are available for further processing. Seismic stations contributing to the data utilized in this study are shown in figure \ref{map:stationMap}.

@@

\chapter{Results} \label{cha:results}

@@

\begin{table}
  \begin{tabular}{ l l l l l l l }
    \cline{2-7}
    & \multicolumn{2}{ c }{ZK} & \multicolumn{3}{ c }{Crust 1.0} & \multicolumn{1}{ c }{Active Source} \\
    \hline
    Region  & $H (km]$& $R$ &$H [km]$& $R$  &$V_P [km/s]$&$V_P [km/s)$ \\
    \hline
    Canada    & 37.0 & 1.75 & 38.1   & 1.74 &   6.41     & 6.33\\
    Shield    & 39.2 & 1.74 & 38.3   & 1.74 &   6.47     & 6.42\\
    Slave     & 38.2 & 1.74 & 38.0   & 1.73 &   6.47     & 6.44\\
    Churchill & 38.9 & 1.73 & 37.9   & 1.74 &   6.46     & 6.39\\
    Superior  & 39.7 & 1.74 & 38.9   & 1.74 &   6.49     & 6.44\\
    Grenville & 41.7 & 1.78 & 39.7   & 1.74 &   6.50     & 6.48\\
    \hline
  \end{tabular}
  \caption[Regional seismic parameter comparison]{Comparison of seismic velocity and crustal thickness estimates for Canada and selected geological provinces.}
\label{table:regionParameters}

@@

\chapter{Discussion} \label{cha:discussion}

\section{Crustal reflectivity}
Seismic refraction and reflection studies of continental crust have afforded valuable constraints on the processes governing crustal evolution. Early efforts to characterize continental crust invoked a 2-layer, slab model comprising lower velocity upper crust and a higher velocity lower crust (Conrad, 1925; Jeffreys, 1926). Evidence supporting this model was originally presented by Conrad (1925) which cites observations of {\it P} arrivals propagating along a mid-crustal interface similar to but shallower than the deeper crust-mantle interface imaged by Andrija Mohorovi\v{c}i\'{c} in 1909 (Litak and Brown, 1989). Interpretations of seismic data invoking both the crustal slab model and the Conrad Discontinuity, as the mid-crustal interface became known, persisted in the literature but were increasingly criticized as being too simplistic, in the case of the slab model, or, lacking consistent evidence (Litak and Brown, 1989) for the Conrad Discontinuity. Technological improvements in reflection seismology allowed researchers to further argue for a complex heterogeneous crust and propose models wherein smooth random lateral inhomogeneities are superimposed on simple velocity gradient models to reproduce the range of seismic observations (Litak and Brown, 1989; Mereu and Ojo, 1981).

@@

\begin{figure}
 \centering
 \includegraphics[width=\textwidth]{discInterp.eps}
 \caption[Normalized station depth profiles and SVD analysis]{Left, vertical station depth profiles interpolated to a common Moho. Right, normalized stack of depth envelopes, grey line shows mean}
 \label{fig:discInterp}
\end{figure}

@@

We reconcile the teleseismic and active source observations of transparent versus reflective lower crust by considering the scale of vertical heterogeneity to which the two methods are sensitive. Teleseismic body waves include both direct and reflected conversions that can originate from layering with velocity gradients over scale lengths of the order of $\lambda/2$ and $\lambda/4$, respectively, where $\lambda$ is the {\it P}-wavelength (Bostock, 1999). For 1 Hz teleseismic waves, crustal layering with scale lengths of $\sim$ 800-2000 m should be resolvable. Higher frequency active source data (10-40 Hz) employing {\it P}-wave sources are sensitive to finer scale vertical structure with scales of $\sim$ 75-300 m. We infer then that the vertical scales of lower crustal reflectivity are better represented by scale lengths under $\sim$ 800 m yielding a lower crustal structure that is relatively transparent to teleseismic waves but can yield strongly reflective signatures to near-vertical, active-source seismic profiling.  Some authors, for example Mayer et al. (1997) in their study of the Rhine Graben, have employed the term Conrad Discontinuity to refer to an interface or transition between transparent upper crust and reflective lower crust. It is possible in this context that the higher energy levels between 15 and 20 km on the 5th principal component in Figure \ref{fig:discPCA}, are a manifestation of such signals, but that the lower energy levels ($<5\%$) in this mode would indicate that the Conrad Discontinuity so-defined is at best weakly represented within the Canadian crust. This finding is supported by crustal velocity models characterized by a near constant velocity gradient such as the gradual transition from a more silicic upper crust to a more mafic lower crust in shields and platforms observed by Meissner (1986). It is also supported by the lack of evidence for a consistent mid-crustal reflection boundary in {\sc lithoprobe} data (Hammer et al., 2010). The increased transparency of the lower crust would appear to relate to changes in rheology. Studies of crustal strength (Jackson, 2002), deformation (Burov and Watts, 2006; Meissner et al., 2006) and the inferred $\sim$ 100 m lower crustal lamallae reported by Mooney and Brocher (1987) as well as the widely noted reduction in seismicity of the lower crust (Ito, 1990; Meissner and Wever, 1986) would support this contention; though see Deichmann (1992), Handy and Brun (2004) and Simpson (1999) for competing evidence.

\section{Crustal Composition} \label{section:crustalcomp}
Mean crustal composition is another important parameter characterizing continental crust though it is important to distinguish it from actual lithological or mineralogical state of the crust. It serves as a representative, average quantity that affords constraint on crustal structure and evolution in space and time. Early attempts at estimating mean crustal composition were largely performed through extrapolations based on the surface sampling of upper crustal rocks (Rudnick and Gao, 2003). After the plate-tectonic revolution, Taylor (1967, 1977), proposed the `island arc' or `andesite' model which posited that bulk crustal chemical composition is equivalent to the average convergent-margin andesite. Several problems with this theory subsequently emerged; for example, average surficial island arc composition is more basaltic than andesitic, and any crustal growth via basaltic basal accretion will only serve to magnify this compositional contrast. Hence this model cannot account for the intermediate composition of post-Archean crust. Recognition of these shortcomings led to the analysis of data from crustal refraction seismic surveys to interrogate middle and lower crustal levels and place constraints on their composition (Christensen and Mooney, 1995; Rudnick and Fountain, 1995). Unlike these latter studies, which also utilize surface and exposed upper-crust geological information, the present work adopts a bulk crustal seismic analysis along the lines of Smithson et al. (1981) although, whereas these authors utilize mean crustal $V_P$, this study combines $V_P/V_S$ from teleseismic measurements with constraints on $V_P$ from previous active source studies as represented in the Crust 1.0 model.

@@

Whereas Zandt and Ammon present evidence for increasing $V_P/V_S$ with crustal age our estimates decrease in $V_P/V_S$ with crustal age. This decrease however is slight, from 1.76 in the Proterozoic units to 1.74 in the Archean units and signifies no trend in $V_P/V_S$ with crustal age. Similarly, whereas Durrheim and Mooney (1991) argue for thicker (>5km) Proterozoic crust with higher {\it P}-velocity than Archean crust, the crustal thickness data presented in this study reveals no significant difference between Proterozoic crust (38.9 km) and Archean crust (38.0 km). The hypothesis that Proterozoic crust is prone to basaltic underplating, thereby explaining higher velocities and thicker crust, suggested by {\it Durrheim and Mooney} (1991) does not seem to explain the data presented in this study. However, this hypothesis has been persuasively augmented by recent work (Rudnick 1995; Rudnick and Gao, 2003) to now include a range of possible processes that may better explain the data.

\chapter{Conclusions}
